---
output:
  md_document:
    variant: markdown_github
bibliography: refs.bibtex
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```

# BayesAdaptive: Bayesian response-adaptive designs for multi-arm sequential clinical trials

The `BayesAdaptive` package was written by Steffen Ventz and is available from 
[Steffen Ventz R Packages](http://bcb.dfci.harvard.edu/~steffen/software.html).

This repository starts with `BayesAdaptive` version 1.0 and implements some 
tweaks to adapt it to current `R version 3.3.3`. 

## Installation

The `BayesAdaptive` package can be installed from GitHub as follows: 

```{r eval = FALSE}
install.packages("devtools")
devtools::install_github(repo = "stefvanbuuren/BayesAdaptive")
```


## More information

The package contains a vignette that demonstrates the functionality. Also, see
the paper by Ventz et.al. [@ventz2017].

## References



