---
title: "Vignette: Introduction to BayesAdaptive"
author: "Steffen Ventz"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to BayesAdaptive}
  %\VignetteEngine{knitr::rmarkdown}
  %\usepackage[utf8]{inputenc}
---

# Introduction

This  document gives a short introduction to the package **BayesAdaptive**, which can be used to design and 
conduct Bayesian response adaptive (BAR) trials.  In the design phase, the package is used to compute operating characteristics of a BAR design, where the package supports either a  subpopulation finding design (SFD) or a subpopulation stratified design (SSD). 
In the experimental phase, while conducting the trial, the package enables  (a) randomizing patients to treatment arms and (b) checking  early stopping rules for futility/efficacy.

Here, we focus  on the design phase and show how to set up a set of simulations. 
Typically we choose the sample size based on a set of alternative treatment scenarios.

  
To simulate trials  you will have to proceed in three steps: 

(1) Simulate arrival and potential outcome data.

(2) Specify  design options and prior parameter. 

(3) Run simulations using the data generated in 1. under the option specified in 2.










#  Arrival and Outcome Data
In the first step you have to generate patient arrival and potential outcome data. 
The functions **arrival.process()** and **get.outcome()** generate such datasets. Throughout the document we  follow a short  example.  We generate 20 datasets and start by defining a variable that specifies the number of datasets, and a list of labels for  disease types,  treatment arms and  subpopulations. 

```{r}
library(BayesAdaptive)
nr.datasets        =20                                  # number of data-sets
NAME  = list(NAME.D=paste0("disease_", 1:2),            # label for disease types    
             NAME.A=c("control", paste0("agent", 1:3)), # label for arms    
             NAME.M=paste0("module", 1:3))              # label for subpopulations    
```





## Arrival data
Patient arrival data-sets are generated with  **arrival.process()** based on a homogeneous Poisson process.  The function requires a matrix of arrival-rates, **rates.by.disease**, and  total sample sizes, **mutants.by.disease**, where for both matrices the element $(m,d)$ denotes the mean accrual per week and the total accrual of patients with disease $d$ in subpopulation $m$. Instead of **mutants.by.disease** you can also specify 
**mutants.by.module** in which case the total number of patients by subpopulation $N_m =\sum_d N_{d,m}$ is fixed, but the number of patients with disease $d$, $N_{d,m}$, is random. We specify a design with 3 subpopulations and 2 disease types, where for each combination of disease and subpopulation we specify an overall sample size of  $N_{d,m}=100$ patients.
```{r}
(mutants.by.disease = matrix(100, 3,2, dimnames = list(NAME$NAME.M, NAME$NAME.D)) )
```
And we specify the  accrual rates as
```{r}
(rates.by.disease   = matrix( c(1.2, 0.7, 1.0, 0.5, 2.2, 0.9), 3,2, 
                              dimnames = list(NAME$NAME.M, NAME$NAME.D)))
```
Lastly we  generate **nr.datasets=20** patient arrival data-sets with **arrival.process()** where we fix the seed at 123.
```{r}
Arrival = arrival.process(
  nr.datasets        = nr.datasets, 
  seed               = 123, 
  rates.by.disease   = rates.by.disease,  
  mutants.by.disease = mutants.by.disease)
```
The output **Arrival** is a list of arrival datasets, where
each element **Arrival[[n]]** is a matrix with  **sum(mutants.by.disease)** $=2 \times 3  \times 100=600$ rows and 3 columns.
The   n-th row of each matrix **Arrival[[n]][i,]** denotes the  arrival time, the disease and the subpopulation of the i-th patient. For instance, the 10th patient in the first data set enters the trial  at week 1,  belongs to subpopulation 3, and has    disease 1.
```{r}
Arrival[[1]][10:12,]
```
We can quickly check the total number of patients in the data-set with  $(d,m)$ combination of disease and mutation
```{r}
## number of patient by subpopulation and disease
xtabs(~ Disease + Module, Arrival[[1]] )
```
which is 100 as desired.



## Potential Outcome  Data 
We also generate potential outcome data before simulating any trial. 
This speeds up computations when we simulate trials. Moreover using the same outcome data in simulations under different prior/randomization parameters reduces the  Monet-Carlo error when we compare trial results across prior/randomization parameters.

To fix a response-scenario, we specify the probability of response to treatment $a$  for each disease and subpopulation $(d,m)$. Let's consider 3 experimental arms and a standard of care. The  response rate equals 0.3 for the standard of care and for the second/third experimental arms  in population one, and a rate of 0.4 for the same arms in population 2 and 3. Agent 1 has a positive treatment effect of 0.25 in subpopulation 1 and 2.

```{r}
rate       = array(dim = c(2,4,3), dimnames = NAME ); 
rate[,,1]  = .3
rate[,,-1] = .4
rate[,2,1] = .55
rate[,2,2] = .65
rate
```

We then generate **nr.datasets=20** potential outcome datasets with  **get.outcome()**. The function  requires the arrival data  **Arrival** as an input
```{r}
Outcome = get.outcome(
    seed        = 123, 
    rate        = rate,
    Arrival     = Arrival)
```


The output **Outcome** is a list of matrices, where each matrix **Outcome[[i]]** has  one row for each patient and one column for each agent.  The n-th row and a-th column corresponds to a potential outcome of the n-th patient when treated with agent a, conditional on the patient's subpopulation and disease. For example the first two patients in the 1st data-set  would respond to the 4th agent, but  fail to respond to agent 3.

```{r}
Outcome[[1]][1:2,]
```



# Design  Parameter

You can specify several design parameters according to your design needs. We elaborate on the most important parameters, and focus  on restricting treatment arms to a subset of subpopulations and diseases; or choosing  hyper-parameter for the Bayesian hierarchical model; and lastly specifying the parameter of the randomization function.


## Treatment Eligibility
You may explore  treatment arms only for a subset of $(d,m)$ combinations of diseases and subpopulations. This can be done by creating an eligibility array **eligibility.array**, 
where **eligibility.array[d,a,m]** equals  one if  agent a is a treatment option for disease/subpopulation (d,m) and zero otherwise. For a design without an active control the user has to set the first column of the array equal to zero, i.e.  **eligibility.array[,1,] =0**.
As an example, let's say we  want to test 3 experimental arms against a disease-specific  control arm. The second  experimental agent will  be tested  only in the first subpopulation and only for the first disease. All other arms are tested for each $(d,m)$ combination. 

```{r}
eligibility.array         = array(1, dim = c(2,4,3), dimnames=NAME)
eligibility.array[,3,-1]  = 0
eligibility.array[-1,3,1] = 0
eligibility.array
```

## Prior parameters 

Adaptive randomization is currently implemented as a binary Probit-regression model, i.e a  normal cdf link function $g = \Phi$, which relates the linear predictor with the probability of treatment success. The probability  model for the treatment outcome  equals  
$$P(R_i = 1 | D_i=d, M_i=m, A_i = a) = p_{d,m,a} = \Phi(\theta_{d,m,a}).$$

The array $\theta = \{ \theta_{d,m,a} \}$ follows a  Gaussian prior, 
where each $\theta_{d,m,a}$ is  decomposed into independent Gaussian components
$\theta_{d,m,a}= \theta_{d,m,0} +I(a>0) \zeta_{d,a,m}.$
The probability $\Phi(\theta_{d,m,0}) = \Phi(\alpha_d + \alpha_{d,m})$
corresponds to the response probability of the standard-of-care, 
such that $\alpha_d \sim N(\mu_d, W_1)$ and $\alpha_{d,m} \sim N(\mu_{d,m}, W_2)$.
Moreover,  $\zeta_{d,m,a} = \beta_a + \beta_{m,a} + \beta_{d,m,a}$ represents the treatment effect on the inverse normal scale. Here  $\beta_a   \sim N( 0, W_3)$ represents the general treatment effect across subpopulations,    $\beta_{a,m} \sim N( 0, W_4)$ is a subpopulation specific random effect,  and $\beta_{d,a,m}  \sim N( 0, W_5)$ is a disease-specific effect within  subpopulation $m$.

Prior parameters have to be specified with  **Prior.list.Fct()**,  which creates the prior covariance matrix of $\theta$ and  internal object for posterior estimation.  The function requires two input parameters, the eligibility array  and a vector which contains the variance parameter $W_i, i=1, \cdots, 5$ for all five parameters. If no  control arms are specified, i.e. **eligibility.array[,1,]=0**, then the function sets the variance $W_1$ and $W_2$ equal to 0 and  suppresses both $\alpha_d$ and $\alpha_{d,m}$ from the model. As an example, we consider a prior such that $p_{d,m,a} = \Phi( \theta_{d,m,a})$ is marginally uniformly distributed on $[0,1]$.

```{r}
Prior = Prior.list.Fct(
         eligibility.array=eligibility.array, 
         Var.vec          =c(W1=.5,W2=.1,W3=.3,W4=.09,W5=.01))
names(Prior)
### mean and -1 times the inverse covariance matrix 
names(Prior$Prior_list)
```


The function **Plot.prior()** can be used to select  suitable prior parameters. With  option **plot.figure=1**, the function plots a 4-panel figure. Panel (a) shows the marginal prior density of the response probability $p_{d,m,a}, a>0$; panel (b) shows the joined  prior densities $(p_{d,m,a}, p_{d',m,a})$ for $d\neq d', a>0$; panel (c) shows the density of $(p_{d,m,a}, p_{d,m',a})$  for $m\neq m'$;  and panel (d) shows  $(p_{d,m,a}, p_{d',m',a})$ for  $d\neq d', m\neq m'$.

```{r,eval=TRUE, fig.width=9, fig.height=3}
Plot.prior(Var=c(.5,.1,.3,.09,.01), plot.figure=1)
```

With option **plot.figure=2** the function shows the conditional distribution of $(p_{d,m,a}, p_{d',m,a})$ given the control treatment arm $(p_{d,m,0}, p_{d',m,0})$ for distinct disease  $d\neq d'$ in the same subpopulation. 

```{r,eval=TRUE,fig.width=8, fig.height=8}
Plot.prior(Var=c(.5,.1,.3,.09,.01), plot.figure=2)
```

Lastly, the  option **plot.figure=3** shows the conditional distribution  $(p_{d,m,a}, p_{d',m',a})$ for distinct disease $d\neq d'$ in different  subpopulation  $m \neq m'$, given the control treatment arm $(p_{d,m,0}, p_{d',m',0})$ . 

```{r,eval=TRUE,fig.width=8, fig.height=8}
Plot.prior(Var=c(.5,.1,.3,.09,.01), plot.figure=3)
```


## Randomization Parameters
Response-adaptive randomization is implemented through the model 
$$
P[ C_i = a | D_i=d, M_i=m, \Sigma_i ] \propto S_{d,m,a}(i) \exp( 5 ** (N^\star  -  N_{d,m,0}(i) )_+ ) I_{d,m,a}(i).
$$
where $x_+ = x I(x>0)$,  $N^\star$ is a design parameter, and $C_i=a$ is the event of randomizing the $i$-th patient to agent $a$ conditional on  the disease and subpopulation $(D_i=d,M_i=m)$. Here $I_{d,m,a}(i)=1$ if agent $a$ is still active for the combination $(d,m)$ at arrival of the i-th patient and 0 otherwise. 

For trials with a  control arm, the statistics $S_{d,m,a}(i)$ equals
$$S_{d,m,a}(i) \propto P[  p_{d,m,a} >   p_{d,m,0}  |  \Sigma_i]^{h(i, d, m )},$$
 for  $a>0$ 
and for the control arm
$$S_{d,m,0}(i) \propto
\exp  \{  c [ \max_{a>0} N'_{d,m,a}(i)   -  N'_{d,m,0}(i)  ]   \},$$
where $N'_{d,m,a}(i)$ represents the number of patients randomized to agent $a$ with disease/subpopulation $(d,m)$. 
For trials without a control arm $S_{d,m,a}(i)$  equals
$$S_{d,m,a}(i) \propto P[ \cap_{a'\neq a} \{ p_{d,m,a} > p_{d,m,0} \}  |  \Sigma_i]^{h(i, m,d )}.$$ 

The package implements the  function $h(i, d,m)=A*N_{d,m}(i)^b$, where the parameters **A** and **b** can be a scalar and identical for each disease and subpopulation $(d,m)$ or dependent on $(d,m)$.

If the planned total sample size for each combination $(d,m)$ is identical, we recommend choosing **A,b** identical across all combinations. If the planned total sample size  differs across $(d,m)$, then **A,b** should be specified as an array, where  $A[d,a,m], b[d,a,m]$ $a\geq 1$ represents the parameter of $h(i, d,m)=a**N_{d,m}(i)^b$ for the combination  $(d,m)$.
Since we specified a total sample size of  $N_{d,m}=100$  patients for each   disease and subpopulation  (d,m), we  use identical parameter and specify $h$ such that
$h(i,d,m) = 1$ after half of the total number of patient in $(d,m)$ are randomized, i.e.  $N_{d,m}(i)= N_{d,m}/2 = 50$, and $h(i,d,m) = 4$ at the end of the trial when $N_{d,m}(i)= N_{d,m} = 100$.

```{r}
N   =100
b   =-log(4)/log(.5)
A   = 4/N^b
```
```{r, echo=FALSE, fig.width=7, fig.height=3} 
par(mar=c(4,3,0,0))
plot(1:N, A * (1:N)^b,  t="l",  col="blue", xlab="patient i", ylab=expression(h[a,d,m](i)) )
abline(h=1)
abline(h=4)
abline(v=50)
```

The randomization parameters are saved as a list

```{r}
rand.vec =list(a = A, b = b, c = .01, N.star = 10)
```

where in addition to **A** and **b**, we also require a minimum of **N.star=10** patients to be randomized to each $(d,m,a)$ combination.




## Decision boundaries

Early stopping rules for futility and efficacy  are based on stopping such that  the combination $(d,m,a)$ is 
stopped for efficacy if $V'_{d,m,a} \geq b'_{d,m,a}(i)$ or stopped for futility if $V''_{d,m,a} \leq b''_{d,m,a}$.

For the SSD design, $V'_{d,m,a}$ is the z-statistics for binary data, and for the SFD design $V'_{d,m,a}$ is the maximum of the z-statistics over all disease $d$ in $(d,m,a)$. The  futility statistics $V''_{d,m,a}$ is the posterior probability of a positive treatment effect for both designs. The stopping functions equal 
$$b'_{d,m,a}(i) = \lambda'_{d,m,a}  ( 1 + s_1  s_2 ^{\overline{N}_{d,m,a}(i) - N.min} )$$
and
$$b''_{d,m,a}(i)= \lambda''  (1- s_3 ^{ \overline{N}_{d,m,a}(i) } )$$.

For both functions $\overline{N}_{d,m,a}(i) = N_{d,m,a}(i)$ for the SSD design,
wheras   $\overline{N}_{d,m,a}(i) = \sum_d N_{d,m,a}(i)$ for the SFD design. Here we select  efficacy and futility boundary parameters  $\lambda'_{d,m,a}=2, s_1=2,s_3=.8$ and  $\lambda''=.05, s_3=0.8$ and plot both  boundaries   in the following figure 

```{r, echo=FALSE, fig.width=7, fig.height=4}
par(mfrow=c(1,2), mar=c(4,3,1,0))
plot(5:30, 2*(1+2*.8 ^ (5:30-5)),  t="l",  col="blue", ylim=c(0,6),
     main="efficacy bounary", xlab="patient i", ylab=expression(b[a,d,m](i)) )

plot(1:30, .05*(1-.8 ^ (1:30)),  t="l", col="blue", ylim=c(0,.07),
     main="futility bounary", xlab="patient i", ylab=expression(b[a,d,m](i)) )
```

The stopping parameters $s_1,s_2,s_3, \lambda', \lambda'', N_{\min}$ are saved as a list 

```{r}
stopping.rules = list(b.futil      = .05, ## lambda''
                      shape.futi   = .9,  ## s_3 
                      b.effic      = 2,   ## lambda'
                      shape1.effic = 2,   ## s_1
                      shape2.effic = .8,  ## s_2
                      N.min        = 15)
```



# Simulating trials

The function **Simulate.trial()** simulates Bayesian  response-adaptive multi-disease, multi-subpopulation trials, and  requires

(1) the arrival data **Arrival**,

(2) the potential outcome data **Outcome**,

(3) the prior parameterd **Prior**,

(4) the parameters for the response-adaptive randomization algorithm **rand.vec** and,

(5) the list of parameters for the early stopping boundaries **stopping.rules**.

We generated all five objects already. In addition, two more parameters have to be specified, namely **Time.Delay** and **Design**. The variable **Time.Delay** is the time period between the beginning of the treatment and the time when the treatment outcome  is evaluated. The variable **Design** is a vector of two elements, where **Design[1]** represents the design option, with $=1$ for the SFD and $=2$ for the SSD. The second element  **Design[2]** represents the randomization option, with  $=1$ if a standard-of-care control arm exists, $=2$ if a historical estimate of $p_{d,m,0}$ for the standard-of-care exists and should be used for the randomization statistics, and $=3$ if no estimate of $p_{d,m,0}$ exists. As explained above, if **Design[2]=3** then $S_{d,m,a}(i) \propto P[ \cap_{a'\neq a} \{ p_{d,m,a} > p_{d,m,0} \}  |  \Sigma_i]^{h(i, d, m )}$ will be used.  

If you choose to use historical estimates of $p_{d,m,0}$, i.e.  **Design[2]=2**, then you have to specify these estimates as a matrix **p.historical**, where the element **p.historical[d,m]** is the probability of a positive response for the $(d,m)$ disease-subpopulation combination.


We select a SSD with active control agent

```{r}
Time.Delay= 8
Design    = c(2,1)
```

## Output options

The simulation output is specified through the vector  **Check**, which has  6  elements, **c(drop, alloc, resp.pr, stat, stat.all, sim.initial)**, all of which should be **FALSE** or **TRUE**. 

The first  component, **Drop**, specifies whether stopping rules should be applied. The next three arguments **alloc, resp.pr** and **stat** control if patient allocation, the empirical response probabilities and the test statistics at arrival of each patient should be returned. If **Check["stat.all"]** equals **FALSE** then the futility and efficacy statistics for the combination **DAM.check = c(d, a, m)** is monitored during the  trial. If last component **Check["sim.initial"]** is **TRUE** then trials are  simulated only until the arrival of the $i^\star$-th patient, where $i^\star = \min \{i : N_{d,m,a}>N_{\min}$ for some $(d,m,a) \}$. As an example,  we apply  early stopping  rules and save the patient allocation  and the efficacy statistics for all  combinations $(d,a,m)$.
```{r}
Check = c(drop        = TRUE,
          alloc       = TRUE,
          resp.pr     = FALSE,
          stat        = TRUE,
          stat.all    = TRUE,
          sim.initial = FALSE)

DAM.check = c(d=1, a=1, m=1)
```
And  simulate 20 trials with **Simulate.trial()**
```{r, eval=TRUE}
trial.outcome = Simulate.trial(
   seed                = 111,
   ArrivalData         = Arrival,
   ResponseData        = Outcome,
   Time.Delay          = Time.Delay,
   Design              = Design,
   Prior               = Prior,
   rand.vec            = rand.vec,
   stopping.rules      = stopping.rules,
   Check               = Check)
```


The output is a list of trial results, in this case 20 trials, with the output  options specified in **Check**
```{r, eval=TRUE}
## number of trials
length(trial.outcome)
## output for the 20th trial
names(trial.outcome[[20]])
```

The element **trial.outcome[[i]]$status** gives an overview of the trial, we check  the results for the second biomarker-subpopulation
```{r}
trial.outcome[[20]]$status[,,,"module2"]
```
The first slice of the array **status** returns the status of each combination of agents and diseases, where $=-1$ denote that the agent was not eligible for the combination, $=0$ that agent was dropped early for futility, $=1$ not dropped for efficacy or futility; and $=2$ stopped for efficacy. The second slice **closing_time** gives the closing time for  arms that were  stopped  early for futility or efficacy.  Lastly, the slices **efficacy_stat** and **futility_stat** return the efficacy and futility statistics at the end of the trial.

Similarly, the output elements **Resp_RiskPop** and **Resp_Events** return the number of observed outcomes and the number of observed positive outcomes at the end of the trial for each combination $(d,m,a)$ of disease, mutations and arms. 

```{r}
trial.outcome[[20]]$Resp_RiskPop[,,"module2"] 
trial.outcome[[20]]$Resp_Events[,,"module2"]
## response probability
trial.outcome[[20]]$Resp_Events[,,"module2"]/
trial.outcome[[20]]$Resp_RiskPop[,,"module2"] 
```
Since we specified the patient allocation in **Check**,  we can also monitor patient allocation during the trial.
```{r, , fig.width=7, fig.height=4}
Col = c("black", "red", "green", "blue")
par(mfrow=c(1,2), mar=c(4,4,2,0))
plot( c(1,1), c(0,0), xlim=c(0,601), xlab="total number of patients i", ylab="Randomized to (1,a,2)", 
      ylim=c(0,max(trial.outcome[[20]]$alloc[,,,2])), main="disease 1" )
for(a in 1:4)  lines(1:601, trial.outcome[[20]]$alloc[1,a,,2], col=Col[a] )
plot( c(1,1), c(0,0), xlim=c(0,601), xlab="total number of patients i", ylab="Randomized to (2,a,2)", 
      ylim=c(0,max(trial.outcome[[20]]$alloc[,,,2])), main="disease 2" )
for(a in 1:4)  lines(1:601, trial.outcome[[20]]$alloc[2,a,,2], col=Col[a] )
legend("topleft", legend = c("Control", paste("Agent", 1:3)), text.col = Col, bty = "n")
```

